
---
title: "Cleaning alldata"
author: "Geoffrey Broadhead"
date: "3/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###2017 and 2018 datasheets have been combined 
There are still lots of irrlevant entries and issues with eclosion day calculations, date format etc. 
1. Mass specific metabolic rates aren't recorded in a saved file/in this data sheet. Need to add those or recalculate. 

AFTER this: 
1. First step is to remove non-experimental data (collected pupae that didn't go into the experiment), and genetic controls that aren't going to be analyzed here. 
2. Later will remove incomplete entries (e.g. animals that died without eclosion etc.)

Reading in the full set of 2017 and 2018 data
```{r message= FALSE, warning = FALSE }
library(readxl)
alldata<-read_excel("E:/Hahn Lab/CircadianRhag/DataProcessing/alldata.xlsx")
```


Locating incomplete data
```{r}
summary(alldata$treatment) # currently 5540 rows of data, the treatments are "character" and not a factor. 
alldata$treatment<-as.factor(alldata$treatment) #converting
#check class of variable again
summary(alldata$treatment)
```

Of the 5540 rows of data, 1377 are GC and 219 have no treatment assigned
Need to remove these.

Should end up with 1988 Room Temp and 1956 Sim Overwintering = 3944 experimental animals (many will not have complete data, more filtering to come)

```{r}
alldata_removedGCNA<- alldata %>%
  filter(treatment %in% c("RT", "SO"))
summary(alldata_removedGCNA$treatment)
```

Are these 3944 total observations complete data?
Need: 
1. Cohort/Collection 
```{r}
